---
title: "`r format(Sys.time(), '%Y')` Icicle Creek Spring Chinook Salmon Forecast"
author: "Jakub Bednarek, Greg Fraser"
date: "Last updated `r Sys.time()-(7*3600)`"
output: html_document
runtime: shiny
---

*This forecast is generated from publicly available data on [PTAGIS.org](https://www.ptagis.org), [Columbia Basin Research DART](http://www.cbr.washington.edu/dart/query/pitadult_hrt), and from [Mid-Columbia Fish & Wildlife Conservation Office](https://www.fws.gov/leavenworthfisheriescomplex/MidColumbiaFWCO/MCFWCOVisit.cfm)*

Passive integrated transponder (PIT) tags are used throughout the Columbia River basin to track migratory fish behavior. PIT tags are used to monitor post-release performance metrics including run timing, survival, and forecasting salmon returns. Based on tagging rates reported by Leavenworth National Fish Hatchery (LNFH, Table 1), we can expand PIT tag counts at Bonneville Dam to forecast returns to Icicle Creek up to four weeks in advance of their arrival. This advanced notice is used to inform harvest decisions and broodstock needs for LNFH. 

The forecast follows the model $N = (B * C) / P$, where $B$ is the PIT count at Bonneville Dam $C$ is the survival rate between Bonneville Dam and Icicle Creek mouth (conversion rate), and $P$ is the proportion of run completed based on average run timing characteristics (Columbia Basin DART, 2018). The conversion rate is estimated using PIT tag interrogation data at the antenna array at lower Icicle Creek (ICL) and Bonneville Dam (BON). 

```{r,  echo = FALSE, message=FALSE, warning = FALSE}
# Load libraries and import PTAGIS dataset
library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)
library(readxl)
library(lubridate)
library(stringr)
library(ggplot2)
#Import PTAGIS Dataset; Data was queried on PTAGIS for LNFH tagged fish
# migrating up the mainstem, including overshoots at TUF and RRF.

shiny::showNotification("Downloading data from PTAGIS. This is a large dataset and may take some time...", duration = 15)# API request for daily report

GET("https://api.ptagis.org/reporting/reports/likeasturgeon/file/LNFH_PIT_IS.csv",
                write_disk("LNFH_PIT_IS.csv", overwrite = TRUE))
LNFH_IS <- read.csv("LNFH_PIT_IS.csv", as.is = TRUE, fileEncoding = "UTF-16LE")

#Clean and tidy the resulting table
#Convert datatypes
LNFH_IS$Tag <- as.character(LNFH_IS$Tag)
LNFH_IS$Site.Code <- substr(LNFH_IS$Site,1,3)
LNFH_IS$First.Time <- mdy_hms(LNFH_IS$First.Time)
# Select useful variables; rename columns to simplify; 
# mutate new columns "Age", and migration Year

# Group antennas by site. 
sitegroups <- tibble(
  Site.Code = c("B2A", "BO1", "BO2", "BO3", "BO4", "MC1", "MC2", "TD1","TD2","JO1", "JO2","PRA"),
  Group = c("BON", "BON", "BON","BON","BON","MCN","MCN","TDL","TDL","JOD","JOD", "PRA")
)

# Can this step be combined with below to make`site_counts`
LNFH_returns <- LNFH_IS %>%
  select(Tag, Site.Code, Brood.Year, First.Time) %>% #select only necessary columns
  left_join(sitegroups) %>% # join antenna groups key so you can group by site R/T antenna
  mutate(MigrationYear = year(First.Time), 
         Age = MigrationYear - Brood.Year, #calculate fish age
         Site.Code = if_else(!is.na(Group), Group, Site.Code)) %>% #collapse sites with only one antenna
 filter(Age > 2 & Age < 7) # filter on adults (Maybe do this later to get more conversion rates)
 # HOWEVER, if you filter age later, you may get inF values for 2014. Looks like no fish detected
 # at bonneville in 2014 for some reason. See when tagging began. 

#spread Site to get site counts and calculate conversion rate annaully
site_counts <- LNFH_returns %>% 
  count(MigrationYear, Site.Code) %>% 
  spread(Site.Code, n, fill=0) %>% 
  arrange(MigrationYear) %>% 
  filter(MigrationYear > 2011 & MigrationYear < year(Sys.Date())) %>%
  mutate(ConversionRate = (ICL / 0.87) / BON) #0.87 is the ICL efficiency 

#Do we get a confidence interval here? or later? use t.test() to get mean and 95% CI
conversion_rate <- mean(site_counts$ConversionRate) #re-write and don't use mean conversion - be more accurate

## Import Pit ratio and clean
pit_ratios <- read.csv("PITratios.csv") %>%
  mutate(Brood.Year = Release.Year-2) %>%
  select(Brood.Year, Release.Number, X..PIT, PIT.Ratio.Non.Tag.Tag) %>%
  rename(NumberTagged = X..PIT, pit_ratio = PIT.Ratio.Non.Tag.Tag) 
# handle missing data by using release target and 
pit_ratios$Release.Number <- if_else(is.na(pit_ratios$Release.Number), 
                                     true = 1200000, false = pit_ratios$Release.Number)
pit_ratios$NumberTagged <- if_else(is.na(pit_ratios$NumberTagged), 
                                   true = 20000, false = pit_ratios$NumberTagged)
pit_ratios$pit_ratio <- if_else(is.na(pit_ratios$pit_ratio), 
                                      true = pit_ratios$Release.Number/pit_ratios$NumberTagged, 
                                      false =pit_ratios$pit_ratio)
```

```{r message = FALSE, warning = FALSE, echo = FALSE}
current_BONcounts <- LNFH_returns %>% 
  count(MigrationYear, Brood.Year, Age, Site.Code) %>% 
  spread(Site.Code, n, fill=0) %>% 
  select(MigrationYear, Brood.Year, Age, BON) %>%
  filter(MigrationYear == year(Sys.Date())) ## replace 2017 with year(Sys.Date())

current_year <- data.frame(Age = as.integer(c(3,4,5)))
current_year <- current_year %>%
  mutate(Brood.Year = as.integer(year(Sys.Date()))-Age) %>%
  left_join(select(current_BONcounts, Brood.Year, BON), by = "Brood.Year" ) %>%
  left_join(pit_ratios, by = "Brood.Year") %>%
  mutate("Expanded Migration Size" = BON * pit_ratio) %>%
  rename("Brood Year" = Brood.Year, "Bonneville Count" = BON, "Number Released" = Release.Number, "Number Tagged" = NumberTagged, "Ratio Released:Tagged" = pit_ratio)%>%
replace_na(list("Bonneville Count" = 0, "Expanded Migration Size" = 0)) 

include = FALSE
# Run Timing Factorization. Expand the run based on run timing
# Query dart for hatchery Spring Chinook released at icicle River. Interestingly, the average run timing is 2 days later than 10 years ago when this script was written!
migration_filepath <- "hrt_pitadult_2025.csv"
dart <- read.csv(migration_filepath)
dart <- dart[,1:10]

# Convert the dart values to Julian day format
julian_dates <- sapply(dart[2, 3:10], function(x) as.numeric(format(as.Date(x, format="%m/%d"), "%j")))

# Define percent_run values corresponding to the date ranges
percent_values <- c(0, 0.05, 0.10, 0.25, 0.5, 0.75, 0.90, 0.95, 1)

# Get today's Julian day
today_julian <- as.numeric(format(Sys.Date(), "%j"))

# Find which interval today falls into and assign the corresponding percent_run value
percent_run <- percent_values[findInterval(today_julian, julian_dates)]
```
## `r if (sum(current_year$"Expanded Migration Size") * conversion_rate / percent_run  < 1 ){
  "No PIT tags have been detected yet. The first Leavenworth fish usually reach Bonneville dam between March 23 and April 19 (Columbia Basin DART 2025)"
} else {
  paste("Forecast to Icicle Creek:", round(sum(current_year$"Expanded Migration Size") * conversion_rate / percent_run))
}`

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(kableExtra)
kable(current_year, 
      "html", 
      align = "c", 
      caption = "Table 1. 2018 LNFH-origin spring Chinook Salmon migration characterists and migration size estimate.") %>%
  footnote( general = "2021 - 2022 releases data and pit tag ratios not yet published. Release and tagging targets are used.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, position = "left")
```

Based on average run timing characteristics the Icicle Creek spring Chinook run is `r paste(100 * percent_run,"%", sep = "")` complete (Columbia Basin Research DART 2018). However, the average run timing may not closely represent this year's run. Beer (2007) found significant correlation between in-stream conditions (temperature and flow) and run timing. 

The conversion rate (C) represents the average survival of adults between Bonneville Dam and the lower Icicle River PIT antenna (Table 2). The ICL antenna detected has an efficiency of 87%, meaning not all of the salmon are detected. The efficiency is factored into the conversion rate. ICL was installed in 2011 and available data includes 2012 to present. The average conversion rate for these years was `r round(conversion_rate, 2)`.

```{r echo = FALSE, message= FALSE, warning=FALSE}
site_counts %>%
  select(MigrationYear, BON, ICL, ConversionRate) %>%
  kable("html",col.names = c("Migration Year", "BON", "ICL", "Conversion Rate"), align = c("c", "c", "l"), caption = "Table 2. PIT tag counts at Bonneville Dam (BON), Icicle Creek Lower Array (ICL) and resulting conversion rates. (PTAGIS, 2018)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, position = "left")
```

To gauge the strength of this model, we compared PIT tag expansions at Bonneville Dam to actual returns to Icicle Creek (Table 3). Icicle Creek returns are measured by hand counted returns to the hatchery, spawning ground surveys, and creel census of sport and tribal fisheries. These sources are used to reconstruct the total return to Icicle Creek.

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(readxl)
icicle_run_reconstruction <- read_excel(
  "Icicle R Spring Chinook Run Reconstruction_01092018.xlsx", 
  sheet = "Icicle Run ReconstructionUpdate", skip = 3)

ici_run <- icicle_run_reconstruction %>%
  select(Year, Run) %>%
  filter(!is.na(Year)) %>%
  rename(MigrationYear = Year)

ici_run$Run <- round(ici_run$Run, 0)

predicted <- LNFH_returns %>% 
  count(MigrationYear, Brood.Year,Site.Code) %>% 
  spread(Site.Code, n, fill=0) %>% 
  arrange(MigrationYear) %>% 
  filter(MigrationYear > 2011 & MigrationYear < year(Sys.Date())) %>%
  inner_join(pit_ratios, by = "Brood.Year") %>%
  mutate(Forecast = round(BON*pit_ratio*conversion_rate, 0)) %>%
  group_by(MigrationYear) %>%
  summarize(Forecast = sum(Forecast)) %>%
  left_join(ici_run, by = "MigrationYear") %>%
  mutate(Difference = Forecast - Run, "Percent Diff" = round((100 * (Run-Forecast) / Run), 0) )
predicted %>%
kable("html",caption = "Table 3. Comparison of PIT expansion method and actual returns to Icicle Creek.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, position = "left")
```

Our forecast typically under predicts actual returns. On average the model under predicts by 21.8%. This may be due to PIT-tagged fish have lower smolt to adult return rates (Knudsen 2009; McDonald et al. 2003; Prentice 1994). PIT tag loss from juveniles has also been documented at the hatcheries. Although loss rate is low, forecasts could underestimate returns based on pit tag ratios biased by tag loss.  

```{r echo = FALSE, warning=FALSE, message=FALSE}
ggplot(predicted, aes(x = MigrationYear))+
  geom_line(aes(y = Forecast, color = "Forecast")) +
  geom_line(aes(y = Run, color = "Run"))+
  ggtitle("Comparison of Run Forecast and Measured Returns")+
  ylab("Number of Spring Chinook")+
  ylim(0, NA)+
  theme_bw()
  
```


### References
    Beer, W.N., 2007. Run timing of adult Chinook Salmon passing Bonneville 
      Dam on the Columbia River, Seattle: Columbia Basin Research, 
      White Paper, University of Washington.
    
    Knudsen C.M., Johnston M.V., Schroder S.L., Bosch W.J., Fast D.E., 
      Strom C.R., 2011. Effects of Passive Integrated Transponder tags on 
      smolt-to-adult recruit survival, growth, and behavior of hatchery 
      spring Chinook Salmon. North Am. Journal of Fisheries Management 29:658-669, 2009
      
    McDonald, T.L., S.C. Amstrup, and B.F.J. Manly. 2003.Tag loss can bias
      Jolly-Seber capture-recapture esti-mates. Wildlife Society Bulletin 31:814-822.
      
    Muir, H., M. Maxey, C. Parker- Graham, T. Bundy, and M. Cooper 2023. Monitoring and Evaluation 
      of the Leavenworth National Fish Hatchery Spring Chinook Salmon Program, 2022. 
      U.S. Fish and Wildlife Service, Leavenworth WA.
    
    Prentice, E.F., D.J. Maynard, S.L. Downing, D.A. Frost,M. S. Kellett, 
      D. A. Bruland, P. Sparks-McConkey, F.W. Waknitz, R. N. Iwamoto, K. McIntyre, 
      and N.Paasch. 1994. Comparison of long-term effects of PIT tags and CW tags 
      on coho salmon (Oncorhynchus kisutch).Pages 123-137 in A study to determine
      the biological feasibility of a new fish tagging system.
